Spinal Atomic Lambda-Calculus

### 2020-01-16 2 - FoSSaCS 2020 (Submitted)/Reviews/Response.txt

We thank the referees for their diligence, and for their helpful remarks and interesting questions. We will respond to a selection from Review 2 and Review 3.

--- Review 2 ---

*** p.8 free covers
I am not completely sure of the definition of free covers. Does c<x> being in the free covers requires c being a free phantom and x a sequence of free variables? Or might x be bound? What about alpha-conversion in this case?

The first is the case: a cover c<x> is free if both c and all variables in x are free. The cover c<x> is bound by a distributor [ c<x> | [\Gamma] ] where [\Gamma] binds x; hence, c and x are bound simultaneously, or not at all. When bound, both c and x may be alpha-converted independently. We will add a formal definition of free covers to clarify this.

--- Review 3 ---

*** p.3 "The observation that the switch rule ...". Who made this observation?

We are making it, in this paper; we will change the emphasis to make this clear.

*** Is there a precise relation between the sharing in the current paper and abstract machines for lazy evaluation a la Haskell? The comment in the paragraph "These notions give rise to four natural duplication regimes...." is too short and only refers to the big-step natural semantics.

We don't think there is a deeper relation, though we'd have to look more closely to be sure. The point of a lazy abstract machine is to (efficiently) implement _when_ duplication takes place, while full laziness and optimal reduction are about _what_ is duplicated. See also the comments below.

*** Perhaps it is a coincidence but most of the abstract machines for Haskell talk about "spine".

This is indeed an unfortunate coincidence of terminology: the "spine" of a lambda term, as relevant to abstract machines, is not the same concept as the "spine" of an abstraction, as relevant here. The former is the path from the root to the leftmost variable; the latter is the tree formed by the paths from a given abstraction to its bound variables. We will add a footnote to point this out.

*** Also it would appear that an abstract machine in the style of Garcia et al. (Lazy evaluation and delimited control) deal with similar issues. [... What] would be connections between current work and control operators?

At present, we have only a limited picture. Fanny He, in her dissertation [The Atomic Lambda-Mu Calculus, University of Bath, 2018] extends the atomic lambda-calculus with mu-abstraction, which is similarly duplicated atomically. This suggests that the two concepts are orthogonal - which is also what the proof theory would indicate. Nevertheless, there may well be operational similarities waiting to be discovered.

### 2020-02-23 3 - FoSSaCS 2020 (Final)/SALC.tex
submitted & final version sources

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{opendeduction}
\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{amsmath, amssymb}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\let\vec\relax
\usepackage{MnSymbol}

\newtheorem{notation}[definition]{Notation}

@@

\newcommand{\FALC}{\Lambda^{S}_{a}}
\newcommand{\SLC}{\Lambda^{S}}
\newcommand{\WEAK}{\Lambda_{\weaksymbol}}
\newcommand{\fv}[1]{(#1)_{fv}}
\newcommand{\bv}[1]{(#1)_{bv}}
\newcommand{\fp}[1]{(#1)_{fp}}
\newcommand{\bp}[1]{(#1)_{bp}}
\newcommand{\fc}[1]{(#1)_{fc}}
\newcommand{\bc}[1]{(#1)_{bc}}
\newcommand{\set}[1]{ \{ #1 \} }
\newcommand{\abs}[2]{\lambda #1 . #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\fake}[3]{#1 \langle \, #2 \, \rangle . #3}
\newcommand{\share}[3]{#1 [#2 \leftarrow #3]}
\newcommand{\dist}[5]{#1 [ #2 \, \vert \, \fakedist{#4}{#5} \, #3 ]}
\newcommand{\olddist}[5]{#1 [ #2 \twoheadleftarrow \lambda #3 \langle #4 \rangle #5]}
\newcommand{\fakedist}[2]{#1 \langle \, #2 \, \rangle}
\newcommand{\barr}[2]{#1 \, \vert \, #2}
\newcommand{\size}[1]{\vert \, #1 \, \vert}
\newcommand{\vecdist}[2]{\overrightarrow{\fakedist{#1}{#2} \,}}
\newcommand{\sub}[3]{#1 \{ #2 / #3 \}}
\newcommand{\lamsub}[2]{\{ \lambda #1 / \lambda #2 \}}
\newcommand{\psub}[3]{#1 \{ #2 / #3 \}_{b}}
\newcommand{\exor}[3]{#1 \{ \fakedist{#2}{#3} \}_{e}}
\newcommand{\readback}[2]{\llbracket \, #1 \, \rrbracket}
\newcommand{\compile}[1]{\llparenthesis \, #1 \, \rrparenthesis}
\newcommand{\weaksymbol}{\mbox{\tiny $\mathcal{W}$}}
\newcommand{\trans}[1]{\llbracket \, #1 \, \rrbracket}
\newcommand{\tranclos}[1]{\| #1 \|}
\newcommand{\readbackclose}[1]{\llbracket \, #1 \, \rrbracket }
\newcommand{\readbackwmap}[3]{\llbracket \, #1 \, \vert \, #2 \, \vert \, #3  \, \rrbracket }
\newcommand{\readweakwmap}[3]{\llbracket \, #1 \, \vert \, #2 \, \vert \, #3  \, \rrbracket_{\weaksymbol} }
\newcommand{\bindvars}[1]{\parallel#1\parallel}
\newcommand{\compweak}[1]{\llparenthesis \, #1 \, \rrparenthesis^{\weaksymbol}}
\newcommand{\readbackweak}[1]{\lfloor \, #1 \, \rfloor}
\newcommand{\composeweak}[1]{\llbracket \, #1 \, \rrbracket^{\weaksymbol}}
\newcommand{\height}[2]{\mathcal{H}^{#1}(#2)}
\newcommand{\weight}[2]{\mathcal{W}^{#1}(#2)}
\newcommand{\weightvar}[2]{\mathcal{V}^{#1}(#2)}
\newcommand{\weightvarshare}[2]{\mathcal{F}^{#1}(#2)}

@@

\newdimen\arrowsize
\pgfarrowsdeclare{delta}{delta}{
        \arrowsize=0.8pt
        \advance\arrowsize by .5\pgflinewidth
        \pgfarrowsleftextend{-.5\pgflinewidth}
        \pgfarrowsrightextend{2\arrowsize-.5\pgflinewidth}
}{
        \arrowsize=.8pt
        \advance\arrowsize by .5\pgflinewidth
        \pgfsetdash{}{0pt} % do not dash
        \pgfsetmiterjoin        % fix join
        \pgfsetbuttcap  % fix cap
        \pgfpathmoveto{\pgfpoint{2\arrowsize}{0pt}}
        \pgfpathlineto{\pgfpoint{0pt}{\arrowsize}}
        \pgfpathlineto{\pgfpoint{0pt}{-\arrowsize}}
        \pgfpathclose
        \pgfusepathqstroke
}

\pgfarrowsdeclare{triangle}{triangle}{
        \arrowsize=1pt
        \advance\arrowsize by .5\pgflinewidth
        \pgfarrowsleftextend{-\arrowsize-.5\pgflinewidth}
        \pgfarrowsrightextend{-.5\pgflinewidth}
}{
        \arrowsize=1pt
        \advance\arrowsize by .5\pgflinewidth
        \pgfsetdash{}{0pt} % do not dash
        \pgfsetmiterjoin        % fix join
        \pgfsetrectcap  % fix cap
        \pgfpathmoveto{\pgfpointorigin}
        \pgfpathlineto{\pgfpointpolar{150}{\arrowsize}}
        \pgfpathlineto{\pgfpointpolar{210}{\arrowsize}}
        \pgfpathclose
        \pgfusepathqfillstroke
}

% PICTURE PRESETS

\tikzstyle{AL}=[>=triangle,x=5mm,y=5mm]
\tikzstyle{dot}=[circle,fill,inner sep=0pt,minimum size=1pt]

% PICTURE COMPONENTS

% triple dots
\newcommand\ALdots[1]{
        \path (#1) node[dot]{} -- +(-3pt,0) node[dot] {} -- +(3pt,0) node[dot]{};
}

        % diagonal bar (for bundles)
\newcommand\ALbar[1]{
        \draw #1 +(2pt,1pt) -- +(-2pt,-1pt);
}
        % general term box
\newcommand\ALnodeT[4][.4]{
        \node [inner sep=4pt, minimum height=.6cm, minimum width=#1cm, shape=rectangle, style=draw, rounded corners] (#4) at (#3) {$#2$};
        \coordinate (#4_top) at ([above=1pt] #4.north);
        \coordinate (#4_bot) at (#4.south);
        \coordinate (#4_lbot) at ([left=3pt] #4.south);
        \coordinate (#4_rbot) at ([right=3pt] #4.south);
}
        % lambda node
\newcommand\ALnodeL[2]{
        \node [inner sep=2pt] (#2) at (#1) {$\lambda$};
        \coordinate (#2_1) at (#2.north);
        \coordinate (#2_2) at (#2.south);
        \coordinate (#2_3) at (#2.east)
}
        % fake lambda node
\newcommand\ALnodeP[2]{
        \node [inner sep=2pt] (#2) at (#1) {$\lambda$};
        \coordinate (#2_1) at (#2.north);
        \coordinate (#2_2) at (#2.south);
        \coordinate (#2_3) at (#2.east)
}
        % application node
\newcommand\ALnodeA[2]{
        \node [inner sep=2pt] (#2) at (#1) {$@$};
        \coordinate (#2_1) at (#2.north);
        \coordinate (#2_2) at (#2.south);
        \coordinate (#2_3) at (#2.east)
}
        % sharing node
\newcommand\ALnodeS[3][.5]{
        \path (#2) -- +(-#1,0.3) coordinate (#3_left) -- +(0,0.3) coordinate (#3_mid) -- +(#1,0.3) coordinate (#3_right) -- +(0,-0.3) coordinate (#3_tip);
        \draw ([above=1pt] #3_tip) -- ([below=1pt] #3_left) -- ([below=1pt] #3_right) -- cycle
}
        % co-sharing node
\newcommand\ALnodeC[3][.5]{
        \path (#2) -- +(-#1,0.3) coordinate (#3_left) -- +(0,0.3) coordinate (#3_mid) -- +(#1,0.3) coordinate (#3_right) -- +(0,-0.3) coordinate (#3_tip);
        \draw [fill=gray] ([above=1pt] #3_tip) -- ([below=1pt] #3_left) -- ([below=1pt] #3_right) -- cycle
}
        % upside-down co-sharing node
\newcommand\ALnodeCup[3][.5]{
        \path (#2) -- +(-#1,-0.3) coordinate (#3_left) -- +(0,-0.3) coordinate (#3_mid) -- +(#1,-0.3) coordinate (#3_right) -- +(0,0.3) coordinate (#3_tip);
        \draw [fill=gray] ([below=1pt] #3_tip) -- (#3_left) -- (#3_right) -- cycle
}
        % nullary sharing node
\newcommand\ALnodeSnull[2]{
        \path (#1) -- +(-.2,.1) coordinate (#2_left) -- +(.2,.1) coordinate (#2_right) -- +(0,-0.1) coordinate (#2_tip);
        \draw (#2_tip) -- (#2_left) -- (#2_right) -- cycle
}

@@

@@

\begin{document}
%
\title{Spinal Atomic Lambda-Calculus}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Tom Gundersen\inst{1} \and
Willem Heijltjes\inst{2}\and
Michel Parigot\inst{3} \and
David Rhys Sherratt\inst{2,4}}
%
\authorrunning{Gundersen et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Red Hat, Inc.  \\ \email{teg@jklm.no} \\ \and
University of Bath, England, UK \\ \email{w.b.heijltjes@bath.ac.uk}
\\ \and
Institut de Recherche en Informatique Fondamentale, CNRS \&\ Universit\'e de Paris, France\\
\email{michel.parigot@gmail.com} \\
\and Institut f\"{u}r Informatik, Friedrich-Schiller-Universit\"{a}t Jena, Germany \\
\email{david.rhys.sherratt@uni-jena.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We present the spinal atomic $\lambda$-calculus, a typed $\lambda$-calculus with explicit sharing and atomic duplication that achieves spinal full laziness: duplicating only the direct paths between a binder and bound variables is enough for beta reduction to proceed. We show this calculus is the result of a Curry--Howard style interpretation of a deep inference proof system, and prove that it has natural properties with respect to the $\lambda$-calculus: confluence and preservation of strong normalisation.

@@

In the $\lambda$-calculus, a main source of efficiency is \emph{sharing}: multiple use of a single subterm, commonly expressed through graph reduction~\cite{Wadsworth-1971} or explicit substitution~\cite{Abadi-Cardelli-Curien-Levy-1991}. This work, and the \emph{atomic $\lambda$-calculus}~\cite{Gundersen-Heijltjes-Parigot-2013-LICS} on which it builds, is an investigation into sharing as it occurs naturally in intuitionistic \emph{deep-inference} proof theory~\cite{Tiu-2006}.

The atomic $\lambda$-calculus arose as a Curry--Howard interpretation of a deep-inference proof system, in particular of the \emph{distribution} rule below left, a variant of the characteristic \emph{medial} rule~\cite{Brunnler-Tiu-2001,Tiu-2006}. In the term calculus, the corresponding \emph{distributor} construct enables duplication to proceed \emph{atomically}, on individual constructors, in the style of sharing graphs \cite{Lamping-1990}. As a consequence, the natural reduction strategy in the atomic $\lambda$-calculus is \emph{fully lazy}~\cite{Wadsworth-1971,Balabonski-2012}: it duplicates only the minimal part of a term, the \emph{skeleton}, that can be obtained by lifting out subterms as explicit substitutions.%
\footnote{While duplication is atomic \emph{locally}, a duplicated abstraction does not form a redex until also its bound variables have been duplicated; hence duplication becomes fully lazy \emph{globally}.}
\[
	\text{\small Distribution:}
\quad
	\drv{A\rightarrow (B\wedge C);-[d];(A\rightarrow B)\wedge(A\rightarrow C)}
\qquad\qquad
	\text{\small Switch:}
\quad
	\drv{(A\rightarrow B)\wedge C;-[s];A\rightarrow(B\wedge C)}
\]
In this work, we investigate the computational interpretation of another characteristic deep-inference proof rule: the \emph{switch} rule above right~\cite{Tiu-2006}.%
\footnote{The switch rule is an intuitionistic variant of \emph{weak} or \emph{linear distributivity}~\cite{Cockett-Seely-1997} for multiplicative linear logic.}
Our result is the \emph{spinal atomic $\lambda$-calculus}, a term calculus in which the natural strategy is a refined form of full laziness, \emph{spine duplication}. This strategy duplicates only the \emph{spine} of an abstraction: the paths to its bound variables in the syntax tree of the term.

We illustrate these notions in Figure \ref{fig:derivationsbalunbal}, for the example term $\lambda x.\lambda y.((\lambda z.z)y)x$. The \emph{scope} of the abstraction $\lambda x$ is the entire subterm, $\lambda y.((\lambda z.z)y)x$ (which may or may not be taken to include $\lambda x$ itself). Note that with explicit substitution, the scope may grow or shrink by lifting explicit substitutions in or out. The \emph{skeleton}, indicated below, is the term $\lambda x.\lambda y.(wy)x$ where the subterm $\lambda z.z$ is lifted out as an (explicit) substitution $[\lambda z.z/w]$. The \emph{spine} of a term, indicated in the second image, cannot naturally be expressed with explicit substitution, though one can get an impression with \emph{capturing} substitutions: it would be $\lambda x.\lambda y.wx$, with the subterm $(\lambda z.z)y$ extracted by a capturing substitution $[(\lambda z.z)y/w]$. Observe further that the skeleton can be described as the \emph{iterated spine}: it is the smallest subgraph of the syntax tree closed under taking the spine of each abstraction, i.e.\ that contains the spine of every abstraction it contains. The terms `skeleton', `spine', and ìterated spine' have already been introduced in \cite{Balabonski-2012}.
%

@@

These notions give rise to four natural duplication regimes. For a shared abstraction to become available as the function in a $\beta$-redex: \emph{laziness} duplicates its \emph{scope} \cite{Launchbury-1993}; \emph{Full laziness} duplicates its \emph{skeleton} \cite{Wadsworth-1971}; \emph{Spinal full laziness} duplicates its \emph{spine} \cite{blanc2005sharing}; \emph{Optimal reduction} duplicates only the abstraction $\lambda x$ and its bound variables $x$ \cite{Lamping-1990,Asperti-Guerrini-1998}.%
\footnote{Interestingly, Balabonski~\cite{balabonski13} shows that for \emph{weak} reduction (where one does not reduce under an abstraction) full laziness and spinal full laziness are both optimal (in the number of beta-steps required to reach a normal form).}

@@

\paragraph*{Switch and Spine.}

One way to describe the skeleton or the spine of an abstraction within a $\lambda$-term is through explicit end-of-scope markers, as explored by Berkling and Fehr~\cite{Berkling-Fehr-1982}, and more recently by Hendriks and Van Oostrom \cite{Hendriks-VanOostrom-2003}. We use their \emph{adbmal} ($\adbmal$) to illustrate the idea: the constructor $\adbmal x.N$ indicates that the subterm $N$ does not contain occurrences of $x$ (or that any that do occur are not available to a binder $\lambda x$ outside $\adbmal x.N$). The scope of an abstraction thus becomes explicitly indicated in the term. This opens up a distinction between \emph{balanced} and \emph{unbalanced} scopes: whether scopes must be properly nested, or not; for example, in $\lambda x.\lambda y.N$, a subterm $\adbmal y.\adbmal x.M$ is balanced, but $\adbmal x.\adbmal y.M$ is not. With balanced scope, one can indicate the skeleton of an abstraction; with unbalanced scope (which Hendriks and Van Oostrom dismiss) one can indicate the spine. We do so for our example term $\lambda x.\lambda y.((\lambda z.z)y)x$ below. 
%
\[
\text{\small Balanced:}
~{\blue\lambda x. \lambda y. (\adbmal y.({\black\adbmal x.\lambda z.z})y)(\adbmal y.x)}
~
\text{\small Unbalanced:}
~{\red\lambda x. \lambda y. ({\black\adbmal x. (\adbmal y.\lambda z.z)y})(\adbmal y.x)}
\]
%
A closely related approach is \emph{director strings}, introduced by Kennaway and Sleep \cite{Kennaway-Sleep-1988} for combinator reduction and generalized to any reduction strategy by Fern\'{a}ndez, Mackie, and Sinot in \cite{Fernandez-Mackie-Sinot-2005}. The idea is to use nameless abstractions identified by their nesting (as with De Bruijn indices), and make the paths to bound variables explicit by annotating each constructor with a string of \emph{directors}, that outline the paths. The primary aim of these approaches is to eliminate $\alpha$-conversion and to streamline substitution. Consequently, while they can \emph{identify} the spine, they do not readily isolate it for duplication. %, as would be required for spinal full laziness.
%\[
%\text{\small Director strings:}~~\red\lambda.(\lambda.({\black((\lambda.\var)\var)^\dirR}\var)^{\dirR\black\dirL})^\dirR
%\]

The observation, that the \emph{switch} rule in open deduction functions as a proof-theoretic end-of-scope construction, serves as the basis of the present work. However, it does so in a \emph{structural} way: it forces a deconstruction of a proof into readily duplicable parts, which together may form the spine of an abstraction. We illustrate this with two typing derivations for our example shown in Figure \ref{fig:derivationsbalunbal}, where we let $x$ have type $A$ and $y,z$ type $A\imp B$, shortened to $\BA$. For a formal definition of how these derivations are constructed, see the next section.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%

\usepackage{graphicx}
\usepackage{opendeduction}
\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing}
\usepackage{amsmath, amssymb}
\usepackage{stmaryrd}
\usepackage{marvosym}
\usepackage{mathtools}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\let\vec\relax
\usepackage{MnSymbol}

@@

\newcommand{\FALC}{\Lambda^{S}_{a}}
\newcommand{\SLC}{\Lambda^{S}}
\newcommand{\WEAK}{\Lambda_{\weaksymbol}}
\newcommand{\fv}[1]{(#1)_{fv}}
\newcommand{\bv}[1]{(#1)_{bv}}
\newcommand{\fp}[1]{(#1)_{fp}}
\newcommand{\bp}[1]{(#1)_{bp}}
\newcommand{\fc}[1]{(#1)_{fc}}
\newcommand{\bc}[1]{(#1)_{bc}}
\newcommand{\set}[1]{ \{ #1 \} }
\newcommand{\abs}[2]{\lambda #1 . #2}
\newcommand{\app}[2]{#1 \, #2}
\newcommand{\fake}[3]{#1 \langle \, #2 \, \rangle . #3}
\newcommand{\share}[3]{#1 [#2 \leftarrow #3]}
\newcommand{\dist}[5]{#1 [ #2 \, \vert \, \fakedist{#4}{#5} \, #3 ]}
\newcommand{\olddist}[5]{#1 [ #2 \twoheadleftarrow \lambda #3 \langle #4 \rangle #5]}
\newcommand{\fakedist}[2]{#1 \langle \, #2 \, \rangle}
\newcommand{\barr}[2]{#1 \, \vert \, #2}
\newcommand{\size}[1]{\vert \, #1 \, \vert}
%\newcommand{\vecdist}[2]{\overrightarrow{\fakedist{#1}{#2} \,}}
\newcommand\vecdist[2]{\vec{#2}}
\newcommand{\sub}[3]{#1 \{ #2 / #3 \}}
\newcommand{\lamsub}[2]{\{ \lambda #1 / \lambda #2 \}}
\newcommand{\psub}[3]{#1 \{ #2 / #3 \}_{b}}
\newcommand{\exor}[3]{#1 \{ \fakedist{#2}{#3} \}_{e}}
\newcommand{\readback}[2]{\llbracket \, #1 \, \rrbracket}
\newcommand{\compile}[1]{\llparenthesis \, #1 \, \rrparenthesis}
\newcommand{\weaksymbol}{\mbox{\tiny $\mathcal{W}$}}
\newcommand{\trans}[1]{\llbracket \, #1 \, \rrbracket}
\newcommand{\tranclos}[1]{\| #1 \|}
\newcommand{\readbackclose}[1]{\llbracket \, #1 \, \rrbracket }
\newcommand{\readbackwmap}[3]{\llbracket \, #1 \, \vert \, #2 \, \vert \, #3  \, \rrbracket }
\newcommand{\readweakwmap}[3]{\llbracket \, #1 \, \vert \, #2 \, \vert \, #3  \, \rrbracket_{\weaksymbol} }
\newcommand{\bindvars}[1]{\parallel#1\parallel}
\newcommand{\compweak}[1]{\llparenthesis \, #1 \, \rrparenthesis^{\weaksymbol}}
\newcommand{\readbackweak}[1]{\lfloor \, #1 \, \rfloor}
\newcommand{\composeweak}[1]{\llbracket \, #1 \, \rrbracket^{\weaksymbol}}
\newcommand{\height}[2]{\mathcal{H}^{#1}(#2)}
\newcommand{\weight}[2]{\mathcal{W}^{#1}(#2)}
\newcommand{\weightvar}[2]{\mathcal{V}^{#1}(#2)}
\newcommand{\weightvarshare}[2]{\mathcal{F}^{#1}(#2)}

@@

@@

\newcommand*{\QEDB}{\hfill\ensuremath{\square}}

\usepackage{lipsum}

\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

@@

\begin{document}
%
\title{Spinal Atomic Lambda-Calculus}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Tom Gundersen\inst{1} \and
Willem Heijltjes\inst{2}\and
Michel Parigot\inst{3} \and
David Sherratt\inst{4} (\Letter)}
%
\authorrunning{Gundersen et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
        Red Hat, Inc. \\ \email{teg@jklm.no} 
\\ \and	University of Bath, UK \\ \email{w.b.heijltjes@bath.ac.uk}
\\ \and Institut de Recherche en Informatique Fondamentale, CNRS, Universit\'e de Paris \\ \email{parigot@irif.fr}
\\ \and Friedrich-Schiller-Universit\"{a}t Jena, Germany \\ \email{david.rhys.sherratt@uni-jena.de}}
%
\maketitle              % typeset the header of the contribution
%
\vspace{-1cm}
\blfootnote{This work was supported by EPSRC Project EP/R029121/1 \emph{Typed Lambda-Calculi with Sharing and Unsharing} and ANR project 15-CE25-0014 \emph{The Fine Structure of Formal Proof Systems and their Computational Interpretations (FISP)}}

\begin{abstract}
We present the spinal atomic $\lambda$-calculus, a typed $\lambda$-calculus with explicit sharing and atomic duplication that achieves spinal full laziness: duplicating only the direct paths between a binder and bound variables is enough for beta reduction to proceed. We show this calculus is the result of a Curry--Howard style interpretation of a deep-inference proof system, and prove that it has natural properties with respect to the $\lambda$-calculus: confluence and preservation of strong normalisation.

@@

In the $\lambda$-calculus, a main source of efficiency is \emph{sharing}: multiple use of a single subterm, commonly expressed through graph reduction~\cite{Wadsworth-1971} or explicit substitution \cite{Abadi-Cardelli-Curien-Levy-1991}. This work, and the \emph{atomic $\lambda$-calculus}~\cite{Gundersen-Heijltjes-Parigot-2013-LICS} on which it builds, is an investigation into sharing as it occurs naturally in intuitionistic \emph{deep-inference} proof theory~\cite{Tiu-2006}. The atomic $\lambda$-calculus arose as a Curry--Howard interpretation of a deep-inference proof system, in particular of the \emph{distribution} rule given below left, a variant of the characteristic \emph{medial} rule~\cite{Brunnler-Tiu-2001,Tiu-2006}. In the term calculus, the corresponding \emph{distributor} enables duplication to proceed \emph{atomically}, on individual constructors, in the style of sharing graphs \cite{Lamping-1990}. As a consequence, the natural reduction strategy in the atomic $\lambda$-calculus is \emph{fully lazy}~\cite{Wadsworth-1971,Balabonski-2012}: it duplicates only the minimal part of a term, the \emph{skeleton}, that can be obtained by lifting out subterms as explicit substitutions.
%
%\footnote{}
(While duplication is atomic \emph{locally}, a duplicated abstraction does not form a redex until also its bound variables have been duplicated; hence duplication becomes fully lazy \emph{globally}.)

	\[
		\text{\small Distribution:}
	\quad
		\drv{A\rightarrow (B\wedge C);-[d];(A\rightarrow B)\wedge(A\rightarrow C)}
	\qquad\qquad
		\text{\small Switch:}
	\quad
		\drv{(A\rightarrow B)\wedge C;-[s];A\rightarrow(B\wedge C)}
	\]

We investigate the computational interpretation of another characteristic deep-inference proof rule: the \emph{switch} rule above right~\cite{Tiu-2006}.%
\footnote{The switch rule is an intuitionistic variant of \emph{weak} or \emph{linear distributivity}~\cite{Cockett-Seely-1997} for multiplicative linear logic.}
Our result is the \emph{spinal atomic $\lambda$-calculus}, a $\lambda$-calculus with a refined form of full laziness, \emph{spine duplication}. In the terminology of \cite{Balabonski-2012}, this strategy duplicates only the \emph{spine} of an abstraction: the paths to its bound variables in the syntax tree of the term.%
\footnote{There is a clash of (existing) terminology: the \emph{spine of an abstraction}, as we use here, is a different notion from the \emph{spine of a $\lambda$-term}, which is the path from the root to the leftmost variable, as used e.g.\ in head reduction and abstract machines.}

We illustrate these notions in Figure \ref{fig:derivationsbalunbal}, for the example $\lambda x.\lambda y.((\lambda z.z)y)x$. The \emph{scope} of the abstraction $\lambda x$ is the entire subterm, $\lambda y.((\lambda z.z)y)x$ (which may or may not be taken to include $\lambda x$ itself). Note that with explicit substitution, the scope may grow or shrink by lifting explicit substitutions in or out. The \emph{skeleton} is the term $\lambda x.\lambda y.(wy)x$ where the subterm $\lambda z.z$ is lifted out as an (explicit) substitution $[\lambda z.z/w]$. The \emph{spine} of a term, indicated in the second image, cannot naturally be expressed with explicit substitution, though one can get an impression with \emph{capturing} substitutions: it would be $\lambda x.\lambda y.wx$, with the subterm $(\lambda z.z)y$ extracted by a capturing substitution $[(\lambda z.z)y/w]$. Observe that the skeleton can be described as the \emph{iterated spine}: it is the smallest subgraph of the syntax tree closed under taking the spine of each abstraction, i.e.\ that contains the spine of every abstraction it contains.

@@

These notions give rise to four natural duplication regimes. For a shared abstraction to become available as the function in a $\beta$-redex: \emph{laziness} duplicates its \emph{scope} \cite{Launchbury-1993}; \emph{Full laziness} duplicates its \emph{skeleton} \cite{Wadsworth-1971}; \emph{Spinal full laziness} duplicates its \emph{spine} \cite{blanc2005sharing}; \emph{optimal reduction} duplicates only the abstraction $\lambda x$ and its bound variables $x$ \cite{Lamping-1990,Asperti-Guerrini-1998}.%
\footnote{Interestingly, Balabonski~\cite{balabonski13} shows that for \emph{weak} reduction (where one does not reduce under an abstraction) full laziness and spinal full laziness are both optimal (in the number of beta-steps required to reach a normal form).}

@@

\paragraph*{Switch and Spine.}
%
One way to describe the skeleton or the spine of an abstraction within a $\lambda$-term is through explicit end-of-scope markers, as explored by Berkling and Fehr~\cite{Berkling-Fehr-1982}, and more recently by Hendriks and Van Oostrom \cite{Hendriks-VanOostrom-2003}. We use their \emph{adbmal} ($\adbmal$) to illustrate the idea: the constructor $\adbmal x.N$ indicates that the subterm $N$ does not contain occurrences of $x$ (or that any that do occur are not available to a binder $\lambda x$ outside $\adbmal x.N$). The scope of an abstraction thus becomes explicitly indicated in the term. This opens up a distinction between \emph{balanced} and \emph{unbalanced} scopes: whether scopes must be properly nested, or not; for example, in $\lambda x.\lambda y.N$, a subterm $\adbmal y.\adbmal x.M$ is balanced, but $\adbmal x.\adbmal y.M$ is not. With balanced scope, one can indicate the skeleton of an abstraction; with unbalanced scope (which Hendriks and Van Oostrom dismiss) one can indicate the spine. We do so for our example term $\lambda x.\lambda y.((\lambda z.z)y)x$ below. 
%
\[
\begin{array}{l@{\qquad}l}
		\text{Balanced scope/skeleton:}   & {\blue\underline{\lambda x.\lambda y.(\adbmal y.}}(\adbmal x.\lambda z.z){\blue\underline{y)(\adbmal y.x)}} 
\\[5pt]	\text{Unbalanced scope/spine:} &  {\red\underline{\lambda x.\lambda y.}}(\adbmal x.(\adbmal y.\lambda z.z)y){\red\underline{(\adbmal y.x)}}
\end{array}
\]

%\begin{array}{ll}
%  \text{Balanced:} & \text{Unbalanced:} \\
%  {\blue\underline{\lambda x. \lambda y. (\adbmal y.(}}\adbmal x.\lambda z.z{\blue\underline{)y)(\adbmal y.x)}}
%& {\red\underline{\lambda x. \lambda y. (}}\adbmal x. (\adbmal y.\lambda z.z)y{\red\underline{)(\adbmal y.x)}}
%\end{array}
%\]
%
A closely related approach is \emph{director strings}, introduced by Kennaway and Sleep \cite{Kennaway-Sleep-1988} for combinator reduction and generalized to any reduction strategy by Fern\'{a}ndez, Mackie, and Sinot in \cite{Fernandez-Mackie-Sinot-2005}. The idea is to use nameless abstractions identified by their nesting (as with De Bruijn indices), and make the paths to bound variables explicit by annotating each constructor with a string of \emph{directors}, that outline the paths. The primary aim of these approaches is to eliminate $\alpha$-conversion and to streamline substitution. Consequently, while they can \emph{identify} the spine, they do not readily isolate it for duplication. %, as would be required for spinal full laziness.
%\[
%\text{\small Director strings:}~~\red\lambda.(\lambda.({\black((\lambda.\var)\var)^\dirR}\var)^{\dirR\black\dirL})^\dirR
%\]

The present work starts from our observation that the \emph{switch} rule of open deduction functions as a proof-theoretic end-of-scope construction (see \cite{sherratt2019atomic} for details). However, it does so in a \emph{structural} way: it forces a deconstruction of a proof into readily duplicable parts, which together may form the spine of an abstraction. The derivations in Figure~\ref{fig:derivationsbalunbal} demonstrate this, as we will now explain---see the next section for how they are formally constructed.

% We illustrate this with two typing derivations for our example, in Figure \ref{fig:derivationsbalunbal}, where $x$ has type $A$ and $y,z$ type $A\imp B$, shortened to $\BA$. For a formal definition of how these derivations are constructed, see the next section.
